---
title: "An introduction to R"
output: html_notebook
---

In this introduction to R, we will present the main R concepts, by introducing R variables, vectors, matrices, lists, functions and packages.

# Variables

In R, we have two main types of values: numeric values and characters.
```{r}
1
"text"
```

We can save these values in variables, so we can easily access them later.
```{r}
a <- 1
a
```

And re-use them :
```{r}
a + a
```


Exercises:
1) Assign a number of your choice to "b"
```{r}

```

2) Show what's in b
```{r}

```

2) Multiply a by 2
```{r}

```

3) Add 10 to a
```{r}

```

4) Assign your name to an object called name
```{r}

```

# Vectors

Vectors contain several values:
```{r}
b <- 1:10
```

You can concatenate these values with the "c" function:
```{r}
c( b, b)
```

You can also apply mathematical formulas to them:

```{r}
b + b
```

Exercices on vectors:
1) Concatenate a and b
```{r}

```

2) Concatenate your name and a
```{r}

```

3) Add a to b
```{r}

```

# Functions

Many functions exist in R, and greatly simplify your life:
```{r}
(1+2+3+4+5+6+7+8+9+10)/10
mean(b)
```

The head function shows the 6 first elements of any object.
The mean function computes the mean of an object.
The length function returns the length of an object.

```{r}
head(b)
head(b, n=3)
median(b)
length(b)
```

# Matrices

Matrices can be compared to tables: they are two dimensional and can contain numbers, characters...
```{r}
load("../inst/extdata/cyto.RData")
cyto
```

We can check the number of rows and columns in the cyto object:
```{r}
nrow(cyto)
ncol(cyto)
```

# Indexing

A specific column of a matrix can be selected by its number:
```{r}
fcs_a <- cyto[,1]
fcs_a
```

Or the column can also be selected by typing its column name directly:
```{r}
cd45 <- cyto[,"CD45"]
cd45
```

Or we can select a specific row:
```{r}
cell2 <- cyto[2,]
cell2
```

We can also select several columns of a matrix, with the "c" function:
```{r}
cyto[,c(1,2)]
```

Exercises on matrices:

1) Show the 2 first rows of the cyto matrix
```{r}

```

2) Show the FSC-A value of the first cell
```{r}

```

# Lists

Lists in R can be compared to vectors, which can contain various objects:
```{r}
random_numbers <- c(1,2,3,7)
random_letters <- c("A","B","C","Z")
```

```{r}
my_list <- list(my_numbers = random_numbers, 
                my_letters = random_letters)
my_list
```

The different members of a list can then be accessed by using the "$" and typing their name directly:
```{r}
my_list$my_numbers
```

Exercises on lists:
1) Create a list containing your name, and job:
```{r}

```


# Packages

Packages contain sets of useful functions, which can be installed, loaded, and then used in R.
A package needs to be installed only once on your computer:

```{r}
install.packages(Rtsne)
```

But needs to be loaded every time you wish to use it:
```{r}
library(Rtsne)
```

The functions of an installed package can then be used. They often need the user to set parameters:

```{r}
tsne <- Rtsne(cyto)
```

```{r}
tsne <- Rtsne(cyto, perplexity = 2)
tsne$Y
```

You can always ask help to R, if you wish to know how a function works:
```{r}
?Rtsne
```
