---
title: "R Notebook"
output: html_notebook
---

# Variables

In R, we have two main types of values: numeric values and characters.
```{r}
1
"text"
```

We can save these values in variables, so we can easily access them later.
```{r}
a <- 1
a
```

And re-use them :
```{r}
a + a
```


Exercises:
1) Assign a number of your choice to "b"
```{r}

```

2) Multiply a by 2
```{r}

```

3) Add 10 to a
```{r}

```

4) Assign your name to an object called name
```{r}

```

# Vectors

Vectors contain several values:
```{r}
b <- 1:10
```

You can concatenate these values with the "c" function:
```{r}
c( b, b)
```

You can also apply mathematical formulas to them:

```{r}
b + b
```

Exercices on vectors:
1) Concatenate a and b
```{r}

```

2) Concatenate your name and a
```{r}

```

3) Add b to a
```{r}

```

# Functions

Many functions exist in R, and greatly simplify your life:
```{r}
(1+2+3+4+5+6+7+8+9+10)/10
mean(b)
```

The head function shows the 6 first elements of any object.
The mean function computes the mean of an object.
The length function returns the length of an object.

```{r}
head(b)
mean(b)
length(b)
```

# Data frames

Data frames can be compared to tables: they are two dimensional and can contain numbers, characters...
```{r}
load("../inst/extdata/cyto_df.RData")
cyto
```

# Indexing

A specific column of a data frame can be selected:
```{r}
fcs_a <- cyto[,1]
fcs_a
```


Or we can select a specific row:
```{r}
cell2 <- cyto[2,]
cell2
```

# Lists

# Packages

```{r}
install.packages()
library()
```
