---
title: "Part 1 - R introduction"
author: "Helena Todorov and Sofie Van Gassen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Part 1 - R introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this introduction to R, we will present the main R concepts, by introducing 
R variables, vectors, matrices, lists, functions and packages.  
You could also have a look at the following online resources:  
https://www.youtube.com/watch?v=o0Y478jOjGk  
https://www.youtube.com/watch?v=u1r5XTqrCTQ  
https://campus.datacamp.com/  
http://tryr.codeschool.com/  

# Variables

In R, we have two main types of values: numeric values and characters.
```{r}
1
"text"
```

We can save these values in variables, so we can easily access them later.
```{r}
a <- 1
a
```

And re-use them :
```{r}
a + a
```


Exercises:

1) Assign a number of your choice to "b"
```{r}
b <- 7
```

2) Show what's in b
```{r}
b
```

3) Multiply a by 2
```{r}
a * 2
```

4) Add 10 to a
```{r}
a + 10
```

5) Assign your name to an object called name
```{r}
name <- "Helena"
```

# Vectors

Vectors contain several values:
```{r}
b <- 1:10
```

You can concatenate these values with the "c" function:
```{r}
c(b, b)
```

You can also apply mathematical formulas to them:
```{r}
b + b
```

Exercises on vectors:  

6) Concatenate a and b
```{r}
c(a, b)
```

7) Concatenate your name and a
```{r}
c(name, a)
```

8) Add a to b
```{r}
a + b
```

# Functions

Many functions exist in R, and greatly simplify your life:
```{r}
(1 + 2 + 3 + 4 + 5 + 6 + 7 + 8 + 9 + 10)/10
mean(b)
```

The head function shows the 6 first elements of any object.  
The median function computes the median of a vector.  
The length function returns the length of a vector.
```{r}
head(b)
head(b, n=3)
median(b)
length(b)
```

# Matrices

Matrices can be compared to tables: they are two dimensional and can contain 
numbers, characters...
```{r}
load("../inst/extdata/cyto.RData")
cyto
```

We can check the number of rows and columns in the cyto object:
```{r}
nrow(cyto)
ncol(cyto)
```

# Indexing

A specific column of a matrix can be selected by its number:
```{r}
fcs_a <- cyto[, 1]
fcs_a
```

Or the column can also be selected by typing its column name directly:
```{r}
cd45 <- cyto[, "CD45"]
cd45
```

Or we can select a specific row:
```{r}
cell2 <- cyto[2, ]
cell2
```

We can also select several columns of a matrix, with the "c" function:
```{r}
cyto[, c(1, 2)]
```

Exercises on matrices:

9) Show the 2 first rows of the cyto matrix
```{r}
cyto[c(1, 2), ]
```

10) Show the FSC-A value of the first cell
```{r}
cyto[1, "FSC-A"]
```

# Lists

Lists in R can be compared to vectors, which can contain various objects:
```{r}
random_numbers <- c(1, 2, 3, 7)
random_letters <- c("A", "B", "C", "Z")
```

```{r}
my_list <- list(my_numbers = random_numbers, 
                my_letters = random_letters)
my_list
```

The different items of a list can then be accessed by using the "$" and 
typing their name directly:
```{r}
my_list$my_numbers
```

Exercises on lists:

11) Create a list containing your name, and job:
```{r}
my_list <- list(name = name, job = "bioinformatician")
my_list
```


# Packages

Packages contain sets of useful functions, which can be installed, loaded, and 
then used in R. A package needs to be installed only once on your computer:
```{r}
install.packages(Rtsne)
```

But needs to be loaded every time you wish to use it:
```{r}
library(Rtsne)
```

The functions of an installed package can then be used. They often need the user to set parameters:
```{r}
tsne <- Rtsne(cyto)
```

```{r}
tsne <- Rtsne(cyto, perplexity = 2)
tsne$Y
```

You can always ask help to R, if you wish to know how a function works:
```{r}
?Rtsne
```
