---
title: "R Notebook"
output: html_notebook
---

# Setup

We load some previously installed libraries, which offer specific
functionality for flow cytometry data analysis. You can have a look at
http://bioconductor.org/packages/release/BiocViews.html#___FlowCytometry
to see more packages which might be of interest.

```{r}
library(flowCore)
library(flowDensity)
library(FlowSOM)
library(FlowSOMworkshop)
```

We point to where the FlowJo workspace and the fcs file are saved on our 
computer. Try to autocomplete with "Tab" to make sure you can access the files 
and the path is correct.

```{r}
wsp_file <- "../inst/extdata/manualGating.wsp"
fcs_file <- "../inst/extdata/21-10-15_Tube_023.fcs"
```

Exercise on finding files:

1) Change the value of fcs_file to Tube 28, using Tab to find the right filename
```{r}

```

# Reading a FlowJo workspace

We parse how this file is processed in FlowJo. This returns a list with two
elements: "flowFrame" contains all information from the fcs file itself,
while "gates" includes all information regarding the gating of the file.

```{r}
flowjo_res <- parse_flowjo(fcs_file, wsp_file)
```

We put the flowFrame and the gating information in variables with a short name,
so we have to type less.

```{r}
ff <- flowjo_res$flowFrame
gates <- flowjo_res$gates
```

# Inspect the flowFrame

The measurements are saved in a matrix, in which each row represents a cell
and every column represents a marker. 

```{r}
nrow(ff)
```

Notice that these column names correspond to the detector names in your
machine.

```{r}
colnames(ff)
```

We can make use of some functions from the FlowSOM package to easily access the
marker names

```{r}
get_markers(ff, colnames(ff))
```

This also works in reverse
```{r}
get_channels(ff, "CD3")
```

To access the specific measurements, we look at the exprs element of the 
flowframe object. Note that these values have been compensated and transformed!

```{r}
head(ff@exprs)
```

Additionally, all metadata describing the fcs file is also stored in the 
flowframe, in the description element.

```{r}
head(ff@description, n = 20)
```

This is a list, so we can also access specific values of interest.

```{r}
ff@description$`$DATE`
ff@description$`$CYT`
```

Exercises on the flowFrame object:

2) Show the number of columns in the flowFrame
```{r}

```

3) Show the values measured for the 150th event
```{r}

```

4) Show which channel was used to measure CD19
```{r}

```

5) Show the CD19 values measured for the first 6 events
```{r}

```

6) Show the median CD19 value
```{r}

```

7) Show at what time the FCS file began recording
```{r}

```

# Inspect the manual gating 

gates is a matrix in which each row represents a cell and each column represents
a FlowJo gate. It contains TRUE or FALSE indicating whether the cell belongs to
the gate (taking the full gating hierarchy into account).

```{r}
head(gates)
```

As FALSE is internally coded as 0 and TRUE as 1, we can calculate the sum per
column to identify how many cells are present in each gate. These numbers
should be close to the numbers you see in the FlowJo workspace.

```{r}
colSums(gates)
```

If one specific gate is of interest, we can select this one column from the 
matrix. Spelling should be exactly the same as in FlowJo!

```{r}
selection_live <- flowjo_res$gates[, "Live"]
```

Exercise:
8) Show how many cells are selected in selection_live
```{r}

```

We can use this vector of TRUE/FALSE values to select the cells of interest.
When a boolean vector of the same length as the data is given for indexation,
only those are selected where the value is TRUE.
```{r}
ff_live <- ff[selection_live, ]
```

Exercise:
9) Show how many cells are present in the new flowFrame
```{r}

```

We can then use this new flowFrame to plot the data of interest
```{r}
channels_to_plot <- get_channels(ff, c("CD64", "AmCyan-A"))

plotDens(ff_live, 
         channels_to_plot)
```

Exercise:
10) Plot the B cells on CD3 vs CD19
```{r}



```

If we want to get the gating information for a subset of cells, we need to
apply the same indexation on the gating matrix.

```{r}
selection_macrophages_from_live <- gates[selection_live, "Macrophages"]
head(selection_macrophages_from_live, n = 10)
```

If we want to assign a color to each cell based on these FALSE/TRUE labels,
we can change the internal 0/1 to 1/2 by adding one,
and using the result to pick the colors
```{r}
color_macrophages <- c("black", "red")[1 + selection_macrophages_from_live]
head(color_macrophages, n = 10)
```

```{r}
channels_to_plot <- get_channels(ff, c("CD64", "AmCyan-A"))

plotDens(ff_live, 
         channels_to_plot,
         col = color_macrophages)
```

Exercise:
10) Show all live cells on CD3 vs CD19, with the B cells colored blue
```{r}



```
