---
title: "Part 5 - FlowSOM"
author: "Helena Todorov and Sofie Van Gassen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Part 5 - FlowSOM}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this script, we showcase how to use the FlowSOM algorithm.

# Setup

Load some previously installed libraries

```{r}
library(flowCore)
library(FlowSOM)
library(FlowSOMworkshop)
```

list.files is a function which list all files in a given directory. A pattern 
can be specified so only the files containing this pattern are selected. 
If you are interested in defining very detailed patterns, have a look at 
"R regular expressions".

```{r}
files <- list.files("../inst/extdata", 
                    pattern = ".fcs")
files
```

We parse all these files, which were manually gated up front. The resulting
data object is a list which contains a flowSet ( = a collection of flowframes)
and a list with the manual gating matrices. Note that all this information is
loaded in the RAM memory of your computer, if you are working with big datasets
this would need to be split up!

```{r}
data <- parse_flowjo(files,
                     "../inst/extdata/manualGating.wsp")
```

We only want to work on the live single cells.

```{r}
data <- gating_subset(data, "Live")
```

We store the first flowframe in a separate variable, for easy access later on

```{r}
ff <- data$flowSet[[1]]
```

While we read information of all gates in the gating hierarchy, some of them
are only used as intermediate steps to finally recover the correct subsets of
cells. We identify all gates which are really of interest to label the cells,
and can then use this information to assign 1 final label to each cell.

```{r}
cell_types <- c("Macrophages", "B cells", "NK cells", "NK T cells",
                "DCs", "Neutrophils", "Basophils", "T cells")

manual_labels <- manual_vector(data$gates,
                               cell_types)
```

# FlowSOM

Now we can run the FlowSOM algorithm.  
We need to pass the data, which channels we want to use, whether we want to 
scale the data (if TRUE, each column is tranformed to a mean of 0 and a sd of 
1), the x- and y-dimensions of our grid (resulting in x * y clusters), the
number of metaclusters (which could be interpreted as the actual populations)
and a seed, which allows to reproduce the exact same random decisions.
One line is added to remove the channel names from the labeling in the plots.

```{r}
channels_of_interest <-  colnames(ff)[c(8:15, 17:19)]
print(get_markers(ff, channels_of_interest))
```

```{r}
fsom <- FlowSOM(data$flowSet,
                colsToUse = channels_of_interest,
                scale = FALSE,
                xdim = 10, ydim = 10,
                nClus = 10,
                seed = 1)
fsom$FlowSOM$prettyColnames <- gsub(" <.*", "", fsom$FlowSOM$prettyColnames)
```

Exercises:

1) Make a fsom49 object with 49 clusters instead of 100 and 8 metaclusters
```{r}
fsom49 <- FlowSOM(data$flowSet,
                  colsToUse = channels_of_interest,
                  scale = FALSE,
                  xdim = 7, ydim = 7,
                  nClus = 8,
                  seed = 1)
```

2) Make a fsom_t_b object for which you only use the CD3 and CD19 markers to 
apply the clustering, using 25 clusters and 4 metaclusters.
```{r}
fsom_t_b <- FlowSOM(data$flowSet,
                    colsToUse = get_channels(ff, c("CD3", "CD19")),
                    scale = FALSE,
                    xdim = 5, ydim = 5,
                    nClus = 4,
                    seed = 1)
```

# Plotting the clustering result

The FlowSOM function returns a list with two element. The first element is 
called FlowSOM and contains all data and information about the first clustering
step. The second element contains the metaclustering result.

We can plot the first clustering by calling the PlotStars function.

```{r fig.width = 10, asp.ratio = 1}
PlotStars(fsom$FlowSOM)
```

In the original FlowSOM object, the size of the node depends on the number of
cells assigned to it. We make an adapted version in which the node size is reset
to a value of 12 everywhere, for an easier view of the data.

```{r fig.width = 10, asp.ratio = 1}
fsom_equal_sizes <- UpdateNodeSize(fsom$FlowSOM, reset = TRUE, maxNodeSize = 12)
PlotStars(fsom_equal_sizes)
```

Exercises:

3) Plot the trees for your fsom49 object

```{r fig.width = 10, asp.ratio = 1}
PlotStars(fsom49$FlowSOM)
```

4)  With the markers argument, you can specify which channels should be shown.  
Plot the tree for the original fsom object showing only four markers of your 
choice.

```{r  fig.width = 10, asp.ratio = 1}
PlotStars(fsom_equal_sizes,
          markers = get_channels(ff, c("CD3", "CD19", "MHCII", "CD161")))
```

5) You can choose the colors of the markers by specifying the colorPalette argument.
This needs a colorPalette as input, which you can generate by using something like
grDevices::colorRampPalette(c("color1", "color2", "color3", "color4"))  
Regenerate your previous plot with some colors of your choice.  
You might want to have a look at colorbrewer2.org to pick nice colors.

```{r fig.width = 10, asp.ratio = 1}
PlotStars(fsom_equal_sizes,
          markers = get_channels(ff, c("CD3", "CD19", "MHCII", "CD161")),
          colorPalette = grDevices::colorRampPalette(c("#bd0026", 
                                                       "#74a9cf", 
                                                       "#2b8cbe", 
                                                       "#df65b0")))
```

6) By specifying view = "grid", you can show the original grid instead. Try
this out for your fsom_t_b object.

```{r fig.width = 15}
PlotStars(fsom_t_b$FlowSOM,
          view = "grid")
```

# Plotting the metaclustering result

The metaclustering level often corresponds closer to the population level you
would expect in manual gating. We visualize them using the backgroundvalues
argument, which takes one value for every FlowSOM cluster as input and color 
codes them.

```{r fig.width = 15}
PlotStars(fsom_equal_sizes,
          backgroundValues = fsom$metaclustering)
```

We can also compare this structure with the manual gating results.
We plot pie charts indicating the percentage of cells in each cluster falling
into a specific gate.

```{r fig.width = 15}
PlotPies(fsom$FlowSOM, 
         manual_labels, 
         backgroundValues = fsom$metaclustering)
```

We can also summarise this in numbers:

```{r}
table(GetMetaclusters(fsom),
      manual_labels)
```

and use this information to label the metaclusters.

```{r}
metacluster_names <- label_metaclusters(fsom, manual_labels)

metacluster_names
```

Exercises:

7) Show the metaclustering on the grid

```{r fig.width = 15}
PlotStars(fsom$FlowSOM,
          backgroundValues = fsom$metaclustering,
          view = "grid")
```

8) Show the manual gating result on the fsom49 tree, with and without the
metaclustering coloring

```{r  fig.width = 15}
PlotPies(fsom49$FlowSOM,
         cellTypes = manual_labels,
         backgroundValues = fsom49$metaclustering)

PlotPies(fsom49$FlowSOM,
         cellTypes = manual_labels)
```

# Inspecting the clusters in more detail

To refer to the clusters, it is easier to assign each of them a number.

```{r  fig.width = 10, fig.height = 10}
PlotNumbers(UpdateNodeSize(fsom$FlowSOM, reset = TRUE, maxNodeSize = 0))
```

Show the metacluster numbers on a plot.

```{r  fig.width = 15}
PlotLabels(UpdateNodeSize(fsom$FlowSOM, reset = TRUE, maxNodeSize = 0),
           labels = fsom$metaclustering)
```

Show the metacluster names on a plot.

```{r  fig.width = 15}
PlotLabels(UpdateNodeSize(fsom$FlowSOM, reset = TRUE, maxNodeSize = 0),
           labels = metacluster_names[fsom$metaclustering])
```

We can then plot a heatmap of the MFIs to have an overview in table format

```{r fig.width = 15}
plot_cluster_MFIs(fsom)
```

```{r fig.width = 15}
plot_metacluster_MFIs(fsom, metacluster_names)
```

It might be interesting to check the backgating of a cluster to the 
traditional scatter plots to ensure that things are making sense.
You can use the PlotClusters2D function for this, which takes a FlowSOM object,
two channel names and a vector of cluster numbers as input.

Example with one cluster:

```{r fig.width = 15}
PlotClusters2D(fsom$FlowSOM,
               get_channels(ff, "CD3"),
               get_channels(ff, "CD19"),
               70)
```

Example with multiple clusters, which all belong to one metacluster:
```{r fig.width = 15}
PlotClusters2D(fsom$FlowSOM,
               get_channels(data$flowSet[[1]], "CD3"),
               get_channels(data$flowSet[[1]], "CD19"),
               which(fsom$metaclustering == 6))
```

Exercises:

9) Plot the metacluster MFIs from the fsom_t_b object

```{r}
plot_metacluster_MFIs(fsom_t_b)
```

10) Plot the metacluster corresponding to the CD3+ CD19- cells from the fsom_t_b
clustering (identified in the heatmap from the previous exercise) on CD3 vs CD19

```{r}
PlotClusters2D(fsom_t_b$FlowSOM,
               marker1 = get_channels(ff, "CD3"),
               marker2 = get_channels(ff, "CD19"),
               which(fsom_t_b$metaclustering == 3))
```

11) From the original FlowSOM, plot the cluster which contains the NKT cells on 
CD3 vs CD161
   
```{r fig.width = 10, asp.ratio = 1}
PlotClusters2D(fsom$FlowSOM,
               get_channels(ff, "CD3"),
               get_channels(ff, "CD161"),
               37)
```

# Showing individual samples

We can take a subset of 1 file at the time.

```{r}
file <- "../inst/extdata/21-10-15_Tube_011.fcs"
fsom_tube11 <- FlowSOM_subset(fsom, file)
PlotStars(fsom_tube11$FlowSOM)
```

Exercises:

12) Plot this figure for Tube 28

```{r}
file <- "../inst/extdata/21-10-15_Tube_028.fcs"
fsom_tube28 <- FlowSOM_subset(fsom$FlowSOM, file)
PlotStars(fsom_tube28)
```

# Comparing samples

This visually already gives some first idea of the variation, but of course it
is better to actually compute this in numbers. 

```{r}
pctgs <- get_pctgs(fsom, metacluster_names)

pctgs$pctgs[, c(1:3, 98:100)]

pctgs$pctgs_meta
```

```{r fig.width = 8, fig.height = 10}
plot_pctgs(pctgs$pctgs_meta)
```

We can also use these numbers to apply statistical tests:

```{r}
statistics_res <- compute_wilcox(pctgs$pctgs_meta,
                                 group1 = c("21-10-15_Tube_011.fcs",
                                            "21-10-15_Tube_012.fcs",
                                            "21-10-15_Tube_013.fcs"),
                                 group2 = c("21-10-15_Tube_028.fcs",
                                            "21-10-15_Tube_030.fcs",
                                            "21-10-15_Tube_031.fcs"))

statistics_res
```
 
